---
title: "Utah Invert Data Analysis"
author: "Gordon Gianniny"
date: "2023-05-18"
output: html_document
---

```{r, include = F}
#Loading packages:

library(tidyverse) #For data wrangling
library(lubridate) #For working with dates
library(ggplot2) #Plotting
library(RColorBrewer)
library(patchwork)
library(stringr)
library(vegan)
library(broom)
library(sp)
```

## Overview

This document carries out all analysis for the Utah Aquatic Insect Decline project. Input files are: 

1. **longterm_site_richness.csv**: yearly average richness and abundance data for each site with at least 4 years of data (n = 142 sites)

2. **longterm_comid_richness.csv**: yearly average richness and abundance data for each COMID segment with at least 4 years of data (n = 211 COMIDs)

Sections are: 

I. Initial data visualizations

    A. By Site
    
    B. By COMID

II. Initial statistics

Read in datafiles: 

```{r}
site_inverts <- read.csv("cleaned_data/longterm_site_richness.csv")
comid_inverts <- read.csv("cleaned_data/longterm_comid_richness.csv")
```

## I. Initial data visualizations

### A. By Site: 

Checking distribution: 

```{r}
ggplot(site_inverts, aes(x = richness_xbar))+
  geom_histogram(fill = "Grey", color = "Black", bins = 70)+
  theme_classic()+
  labs(x = "Mean Richness", y = "Count")
```

Data are fairly normally distributed, maybe slightly bimodal. 

What are the general trends in richness over time at each site? Simplified faceted scatterplot + regression line for each site: 

```{r, warning=F}
site.trends <- ggplot(site_inverts, aes(y = richness_xbar, x = yr, color = factor(siteId)))+
  geom_point(size = 0.5)+
  geom_smooth(se = F, method = "lm", size = 0.5)+
  theme_classic()+
  facet_wrap(~factor(siteId), scales = "free")+
  theme(legend.position = "none", 
        strip.text = element_blank(),
        strip.background = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())+
  labs(x = "Year", 
       y = "Mean Richness")
site.trends
```

A few initial observations: 

  * No consistent trend across all sites - some look flat, some increase, some decrease. 
  
  * Seems like most sites with longer-term data (more points) generally have increasing trends. 
  
  * Very clear trends (+ or -) at some sites, others are more scattered. 
  
  * Would be cool to see if there's some other variable (location, elevation, land use, etc) that explains different trends @ different sites (i.e. why some are increasing, others are decreasing)


Save: 

```{r}
ggsave("plots/site_richness_trends.jpg", site.trends, device = "jpeg", height = 8.5, width = 8.5, units = "in", dpi = "retina")
```

### B. By COMID:

Checking distribution: 

```{r}
ggplot(comid_inverts, aes(x = richness_xbar))+
  geom_histogram(fill = "Grey", color = "Black", bins = 70)+
  theme_classic()+
  labs(x = "Mean Richness", y = "Count")
```

Same as above, might be slightly more bimodal. 

What are the general trends in richness over time for each COMID? Simplified faceted scatterplot + regression line for each COMID: 

```{r, warning=F}
comid.trends <- ggplot(comid_inverts, aes(y = richness_xbar, x = yr, color = factor(COMID)))+
  geom_point(size = 0.5)+
  geom_smooth(se = F, method = "lm", size = 0.5)+
  theme_classic()+
  facet_wrap(~factor(COMID), scales = "free")+
  theme(legend.position = "none", 
        strip.text = element_blank(),
        strip.background = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())+
  labs(x = "Year", 
       y = "Mean Richness")
comid.trends
```

Save: 

```{r}
ggsave("plots/comid_richness_trends.jpg", comid.trends, device = "jpeg", height = 8.5, width = 8.5, units = "in", dpi = "retina")
```




