---
title: "Utah Invert Data Analysis"
author: "Gordon Gianniny"
date: "2023-05-18"
output: html_document
---

```{r, include = F}
#Loading packages:

library(tidyverse) #For data wrangling
library(lubridate) #For working with dates
library(ggplot2) #Plotting
library(RColorBrewer)
library(patchwork)
library(stringr)
library(vegan)
library(broom)
library(sp)
```

## Overview

This document carries out all analysis for the Utah Aquatic Insect Decline project. Input files are: 

1. **longterm_site_richness.csv**: yearly average richness and abundance data for each site with at least 4 years of data (n = 142 sites)

2. **longterm_comid_richness.csv**: yearly average richness and abundance data for each COMID segment with at least 4 years of data (n = 211 COMIDs)

Sections are: 

I. Initial data visualizations

    A. By Site
    
    B. By COMID

II. Initial statistics

Read in datafiles: 

```{r}
site_inverts <- read.csv("cleaned_data/longterm_site_richness.csv")
comid_inverts <- read.csv("cleaned_data/longterm_comid_richness.csv")
```

## I. Initial data visualizations

### A. By Site: 

Checking distribution: 

```{r}
ggplot(site_inverts, aes(x = richness_xbar))+
  geom_histogram(fill = "Grey", color = "Black", bins = 70)+
  theme_classic()+
  labs(x = "Mean Richness", y = "Count")
```

Data are fairly normally distributed, maybe slightly bimodal. 

### B. By COMID:





